---
title: ""
execute:
  echo: false
format:
  html:
    theme: cosmo
    mainfont: proxima-nova, sans-serif
    fig-width: 8
    fig-height: 4
    fig-dpi: 400
    page-layout: full
    linestretch: 1.0
---

```{r}
#| include: false

library(tidyverse)
library(lubridate)
library(tsibble)
library(zoo)
library(hrbrthemes)
library(cowplot)
library(ggtext)
library(here)
# necessary for cowplot
set_null_device("png")
```

```{r}
# parameters
# current fed mw
current_fed_mw <- 7.25

# inflate to current_cpi_date dollars
current_cpi_date <- "2023m7" 

# for figure we usually label current date as month after cpi date
current_cpi_date_label <- "Aug. 2023"
```


```{r}
fed_mw <- here("data", "raw", "fed_min_wage.csv") %>% 
  read_csv(show_col_types = FALSE) %>% 
  mutate(date = yearmonth(mdy(date))) %>% 
  add_row(date = yearmonth(current_cpi_date), fed_min_wage = current_fed_mw) %>% 
  as_tsibble(index = date) %>% 
  fill_gaps() %>% 
  arrange(date) %>% 
  fill(fed_min_wage, .direction = "down") %>% 
  select(date, nominal_fed_mw = fed_min_wage) 

# created by create_cpi.R
cpi <- here("data", "processed", "cpi_u_rs_extended.csv") %>% 
  read_csv(show_col_types = FALSE) %>% 
  mutate(date = yearmonth(date)) 

cpi_base <- cpi %>% 
  filter(date == yearmonth(current_cpi_date)) %>% 
  pull(cpi_u_rs_sa_extended)

real_fed_mw <- fed_mw %>% 
  full_join(cpi, by = "date") %>% 
  mutate(real_fed_mw = nominal_fed_mw * cpi_base / cpi_u_rs_sa_extended) %>% 
  filter(date >= yearmonth("1938m10")) %>% 
  select(date, real_fed_mw, nominal_fed_mw) 

date_breaks <- tribble(
  ~date, ~date_label, ~date_hjust, ~date_description,
  ym("1938m10"), "Oct. 1938", 0, NA,
  ym("1968m2"), "Feb. 1968", 0.5, "Peak value", 
  ym("2009m7"), "July 2009", 0.5, "Last increase",
  ym(current_cpi_date), current_cpi_date_label, 1, "Today"
)

line_graph_data <- real_fed_mw %>% 
  # easier to graph with Date data
  mutate(date = ym(date)) %>% 
  full_join(date_breaks, by = "date") %>% 
  mutate(
    date_break = if_else(!is.na(date_label), date, NA),
    date_value = if_else(!is.na(date_label), real_fed_mw, NA),
    date_value = round(date_value, 2),
    date_value_string = if_else(
      !is.na(date_value),
      scales::label_dollar(accuracy = 0.01)(date_value),
      ""
    )
  )

bar_graph_data <- line_graph_data %>% 
  filter(!is.na(date_description)) %>% 
  mutate(date_year = year(date)) %>% 
  mutate(date_description = case_when(
    date_year == 1968 ~ "Peak value of minimum wage",
    date_year == 2009 ~ "Last minimum wage increase",
    date_year == year(yearmonth(current_cpi_date)) ~ "Today"
  )) %>% 
  mutate(date_value_string = case_when(
    date_year == 1968 ~ paste0(date_value_string, "/hr"),
    TRUE ~ date_value_string
  )) %>% 
  mutate(
    annual_value_string = prettyNum(round(date_value*2080), big.mark=","),
    date_value_string = paste0("**", date_value_string, "**"),
    date_value_string = paste0(date_value_string, " ($", annual_value_string, ")")
  ) %>% 
  arrange(date) %>% 
  mutate(display_order = 4-row_number())



value_peak <- bar_graph_data %>% 
  filter(year(date) == 1968) %>% 
  pull(real_fed_mw)
decline_since_peak <- abs(current_fed_mw / value_peak - 1) * 100
decline_since_peak <- paste0(round(decline_since_peak), "%")

value_last <- bar_graph_data %>% 
  filter(year(date) == 2009) %>% 
  pull(real_fed_mw)
decline_since_last <- abs(current_fed_mw / value_last - 1) * 100
decline_since_last <- paste0(round(decline_since_last), "%")

my_theme <- function(){
  theme_ipsum() +
  theme(
        text = element_text(family = "Proxima Nova Rg"),
        plot.margin = margin(
          t = 5, 
          r = 5,
          b = 5, 
          l = 0, 
          unit = "pt"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none"
  )
}

real_fed_mw %>% 
  mutate(
    across(matches("mw"), ~ round(.x, 2)),
    month = month(date),
    year = year(date)
  ) %>% 
  select(date, year, month, nominal_fed_mw, real_fed_mw) %>% 
  write_csv("real_federal_minimum_wage.csv")
```


```{r}
p1 <- line_graph_data %>% 
  ggplot() +
  geom_segment(
    aes(
      x = date_break, 
      xend = date_break, 
      y = 3.75, 
      yend = 14
    ),
    color = "grey80",
    linetype = "dashed",
    na.rm = TRUE
  ) +
  geom_line(
    aes(x = date, y = real_fed_mw), 
    data = filter(line_graph_data, date <= ym("2009m7")),
    linewidth = 1.5,
    color = "#1273af",
    lineend = "round"
  ) +
  geom_line(
    aes(x = date, y = real_fed_mw), 
    data = filter(line_graph_data, date >= ym("2009m7")),
    linewidth = 1.5,
    color = "#c01f41",
    lineend = "round"
  ) +
  my_theme() +
  scale_x_date(
    breaks = NULL,
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = c(6, 8, 10, 12),
    minor_breaks = NULL,
    labels = scales::label_dollar()
  ) +
  coord_cartesian(
    xlim = c(ym("1938m1"), my(current_cpi_date_label)), 
    ylim = c(3, 13.75),
    expand = FALSE
  ) +
  geom_text(
    aes(
      x = date, 
      y = 3.5, 
      label = date_label, 
      hjust = date_hjust
    ),
    na.rm = TRUE,
    color = "grey40",
    family = "Proxima Nova Rg"
  ) +
  geom_text(
    aes(
      x = date, 
      y = 13.5, 
      label = date_value_string, 
    ),
    hjust = 1,
    nudge_x = -300,
    na.rm = TRUE,
    color = "#2b2b2b",
    family = "Proxima Nova Rg",
    fontface = "bold"
  ) +
  geom_text(
    aes(
      x = date, 
      y = 13, 
      label = date_description
    ),
    hjust = 1,
    nudge_x = -300,
    na.rm = TRUE,
    color = "grey40",
    family = "Proxima Nova Rg"
  ) +
  annotate(
    "text", 
    x = ym("1938m10") + 300, 
    y = 13.5, 
    label = "Minimum wage",
    color = "grey40",
    family = "Proxima Nova Rg",
    hjust = 0
  ) + 
  annotate(
    "text", 
    x = ym("1938m10") + 300, 
    y = 13, 
    label = "established",
    color = "grey40",
    family = "Proxima Nova Rg",
    hjust = 0
  )
```

```{r}
p2 <- bar_graph_data %>% 
  ggplot(aes(x = display_order, y = date_value, fill = factor(display_order))) +
  geom_col(width = 0.5) + 
  my_theme() +
  theme(plot.margin = margin(t = 20, r = 5, l = 25, b = 5, unit = "pt")) +
  coord_flip() +
  scale_fill_manual(values=c("#c01f41", "#1273af", "#1273af")) +
  scale_x_continuous (
    breaks = NULL,
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    expand = c(0,0),
    breaks = NULL,
    minor_breaks = NULL
  ) +
  geom_richtext(
    aes(
      x = display_order, 
      y = date_value, 
      label = date_value_string, 
    ),
    size = 5,
    hjust = 1,
    vjust = 0.57,
    nudge_y = -0.1,
    color = "#ffffff",
    family = "Proxima Nova Rg",
    label.color = NA
  ) +
  geom_text(
    aes(
      x = display_order, 
      y = 0, 
      label = date_year, 
    ),
    size = 5,
    hjust = 0,
    vjust = 0.57,
    nudge_x = 0.45,
    family = "Proxima Nova Rg",
    fontface = "bold"
  ) +
  geom_text(
    aes(
      x = display_order, 
      y = 0, 
      label = date_description, 
    ),
    size = 5,
    hjust = 0,
    vjust = 0.57,
    nudge_x = 0.45,
    nudge_y = 0.85,
    family = "Proxima Nova Rg",
    color = "grey40"
  )
```


### After the longest period in history without an increase, the federal minimum wage today is worth `r decline_since_last` less than 13 years ago--and `r decline_since_peak` less than in 1968 
#### <b>Real value of the minimum wage (adjusted for inflation)</b>

```{r fig.height=7, fig.width=8, warning=FALSE}
plots <- plot_grid(p1, p2, nrow = 2, rel_heights = c(3,2))
plots
```
<small>All values in `r format(yearmonth(current_cpi_date), format = "%B %Y")` dollars, adjusted using the CPI-U in 2023 chained to the CPI-U-RS (1978–2022) and CPI-U-X1 (1967–1977) and CPI-U (1966 and before). Calculations by Ben Zipperer, Economic Policy Institute. Download the [data](real_federal_minimum_wage.csv) or view the [code](https://github.com/Economic/real_minimum_wage) that produces this figure.</small>